Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-18 ieee_paper.tex
Small corrections.

Common improvement techniques such as register blocking, bandwidth
reduction (matrix reordering~\cite{Cuthill69}), partitioning to fit in
cache or TLB~\cite{Nishtala07,Temam:1992:CBS:147877.148091,conf/ppsc/Toledo97},
unrolling~\cite{Mellor-Crummey04} have impacts which are very
dependent on the matrix and overall do not lead to dramatic
improvement. (The state-of-the-art techniques in OSKI~\cite{Vuduc05}
provide some useful yet limited improvements). Register blocking~\cite{conf/ppsc/Toledo97} does
not apply well to most the matrix at hand (despite it can be applied
with virtually no overhead in all cases thanks to compressed
representation~\cite{Buluc11}). There is about 8 bytes of the matrix to move
in per non zero (assuming single precision); each nonzero requires two
floating point operations leading to a flop-to-byte ratio of at most
$\frac{1}{4}$. This limits the obtained performance to at most a
quarter of the bandwidth of the architecture wasting a lot of
potentially useful cycles. The commonly used techniques are mostly
designed to reach that bound rather than overcome it.

@@

where $f$ is the source vector of function values and $g$ is the
resulting vector of derivative values. The computation of a single
derivative has been reduced to a SpMV, where each row has $n_z$
nonzeros. In
practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are similar to what is used in
finite-element codes. Notice that the non zero elements of the matrix
correspond to a relation of nearest neighbor. This relationship is not
necessarily symmetric as shown in Figure~\ref{fig:rbf_stencils} which
means that the matrix sparsity pattern might not be symmetric either.

@@

In term of memory layout, we choose a format similar to the ELLPACK
format to store the matrix since the number of non zero per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. For each row it consists of $n_z$
consecutive entries which give in which column is the non zero. The
other array is the {\tt data} array which gives the values of the non
zero in the four matrices. The values of the matrices are interleaved
so that the value $A^l_{col\_id[k]}$ of the $k$th non zero of the
$l$th matrix is at index $(k-1)*4+l-1$. The vectors {\tt x} are also
similarly interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly
interleaved. This is schematically shown in Figure~\ref{fig:mat_struct}.

@@

\begin{figure}[bth]
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of non-zero\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the matrice(s) in bytes\\
        $s_x$ & size of the {\tt x} vector(s) in bytes\\
        $s_y$ & size of the {\tt y} vector(s) in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred  \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred  \\
        $O$     & number of floating point operations \\
        $I_b$   & maximal computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notations}
  \label{tab:not}
\end{figure}

@@

The matrix is composed of $n_r$ rows and columns with $n_z$ non-zeros
per row leading to a total of $$n_c = n_r n_z$$ non-zero entries in
the matrix. Each of these non-zero entries has one index of size $b_i$
and $n_m$ values of size $b_x$. The matrices have a total size
of $$s_M = n_c (b_i + b_x n_m) = n_r n_z (b_i + b_x n_m)$$

@@

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of non-zeros in the
matrix. Note however that most architectures cannot read memory a
single byte at a time. Instead, a minimum number of bytes, equal to
the size of a cacheline $cl$, are transfered at once.  When there are
multiple vectors, each non-zero element uses $n_v$ consecutive
entries. The worst case number of bytes read and transfered are
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

In SpMV, each non-zero of the matrix requires two floating point
operations: one for performing the multiplication and one for
accumulating the result row-wise. Here we are dealing with $n_v n_m$
simultaneous SpMVs and the number of floating point operations is
$$O = 2 n_v n_m n_c = 2 n_v n_m n_z n_r$$

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show in Section~\ref{sec:expe}
that an implementation relying on compiler vectorization does not lead
to desirable performance, which confirms the results
of~\cite{Saule13-ARXIV}.) The registers in the MIC architecture are
512 bits wide and can store 16 floats or integers. (In comparison to
OpenCL or CUDA, one can think of these registers as a warp.) The end
goal is to be able to load and format the non zero values $a,b,c,d$
and the vector entries $p,q,r,s$ in two vector registers to be able to
apply Fused Multiply-Add on them, which is depicted in
Figure~\ref{fig:tensor_product}. Bringing the data in the vector
register as depicted in the figure is the difficult part that we now
explain. A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
using an OpenMP construct.

@@

All the operations that loads data from the memory into a register
allows to transfer 512 bits, so 16 floats at a time. A single non zero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we need to load at once the data for 4 non zeroes in the matrix,
representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of non zeroes on a row in our application is
always a multiple of 4, if it was, we would need to add some explicit
zeroes in the data structure to pad it.)

The values from the non zero element in the matrix are the simplest to
load. One can load the 16 floating point values for 4 non zeroes of
the matrix at once. The 4 floats coming from each matrix four 4
consecutive non zeroes are naturally grouped within each lane of the
SIMD vector. One can use a swizzle operation to broadcast the 4
elements of a single matrix to take the whole vector. (The broadcast of 
$A_h^1,A_h^2,A_h^3,A_h^4$ is shown in Figure~\ref{code:mat_mul}.)

Loading the entries from the {\tt x} vector is a little bit more
complicated. First the columns pointers of 4 different non zeroes are
loaded in a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to unpack for the
  LSB and MSB. Both are mandatory even if one is known not to be
  useful according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
replicated so that each fill a vector lane using a swizzle
operation. Then each value is multiplied by 4 (because there are 4
vectors) and the offsets $(3,2,1,0)$ is added each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a SIMD register where each lane is the
four vector entries for a non zero element. Using a broadcast
permutation, one can replicate an entry of the four vectors accross
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the non zeroes of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize traffic on the memory bus.

@@

To perform our experiments and analysis, we use multiple kinds of
matrices which are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have non zero elements in the first 32 columns of
the matrix. That way we ensure the simplest form of multiplication,
the amount of elements from the {\tt x} vector data transfered from
the memory will be minimum. The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 non zeroes per row
centered around the diagonal and it represents the ideal case for most
application relying on sparse matrix vector multiplication. The Random
matrices (Figure~\ref{fig:compact}) see their non zero element
randomly (uniformly) distributed in the matrix, they represent the
worst case scenario for cache based architecture where the cache
reutilisation is the lowest from one row to the next.

@@

In the previous sections, we have explored in the practical
implementation on a MIC architecture of multiple derivative operators
acting on multiple vectors within the context of RBF-RD. Each
derivative has an associated sparse matrix with a fixed number of
non-zeros on each row. While computing a single derivative of multiple
functions is rather common (vectorization occurs over the vectors), we
accelerate the algorithm further by considering multiple derivatives
with corresponding matrices with identical adjacency graphs. We
specialize to four matrices and four derivatives, with 16 outputs,
computed as a sum of outer products.

>>>>

Common improvement techniques such as register blocking, bandwidth
reduction (matrix reordering~\cite{Cuthill69}), partitioning to fit in
cache or TLB~\cite{Nishtala07,Temam:1992:CBS:147877.148091,conf/ppsc/Toledo97},
unrolling~\cite{Mellor-Crummey04} have impacts which are very
dependent on the matrix and overall do not lead to dramatic
improvement. (The state-of-the-art techniques in OSKI~\cite{Vuduc05}
provide some useful yet limited improvements). Register blocking~\cite{conf/ppsc/Toledo97} does
not apply well to most the matrix at hand (despite it can be applied
with virtually no overhead in all cases thanks to compressed
representation~\cite{Buluc11}). There is about 8 bytes of the matrix to move
in per nonzero (assuming single precision); each nonzero requires two
floating point operations leading to a flop-to-byte ratio of at most
$\frac{1}{4}$. This limits the obtained performance to at most a
quarter of the bandwidth of the architecture wasting a lot of
potentially useful cycles. The commonly used techniques are mostly
designed to reach that bound rather than overcome it.

@@

where $f$ is the source vector of function values and $g$ is the
resulting vector of derivative values. The computation of a single
derivative has been reduced to a SpMV, where each row has $n_z$
nonzeros. In
practice, $n_z=32$ in two-dimensional flows and 64 or 100 for
three-dimensional flows. These numbers are similar to what is used in
finite-element codes. Notice that the nonzero elements of the matrix
correspond to a relation of nearest neighbor. This relationship is not
necessarily symmetric as shown in Figure~\ref{fig:rbf_stencils} which
means that the matrix sparsity pattern might not be symmetric either.

@@

In term of memory layout, we choose a format similar to the ELLPACK
format to store the matrix since the number of nonzero per row is
constant throughout the matrix. The matrix is given in two arrays. The
first one is the classical {\tt col\_id} array that describes the
sparsity structure of the matrix. For each row it consists of $n_z$
consecutive entries which give in which column is the nonzero. The
other array is the {\tt data} array which gives the values of the non
zero in the four matrices. The values of the matrices are interleaved
so that the value $A^l_{col\_id[k]}$ of the $k$th nonzero of the
$l$th matrix is at index $(k-1)*4+l-1$. The vectors {\tt x} are also
similarly interleaved so that {\tt x}$^l_i$ and {\tt x}$^{l+1}_i$ are
consecutive in memory. The vector {\tt y} is similarly
interleaved. This is schematically shown in Figure~\ref{fig:mat_struct}.

@@

\begin{figure}[bth]
  \begin{center}
    \scalebox{.9}{
      \begin{tabular}{|c|l|}
        %\hline
        %& & \\
        \hline
        $b_i$ & number of bytes per index \\
        $b_x$ & number of bytes per value \\
        $n_z$ & number of nonzeros per row $A$ \\
        $n_r$ & number of column/rows of $A$ \\
        $n_c$ & total number of nonzeros\\
        $n_v$ & number of {\tt x} vectors \\
        $n_m$ & number of matrices \\
        $s_M$ & size of the matrice(s) in bytes\\
        $s_x$ & size of the {\tt x} vector(s) in bytes\\
        $s_y$ & size of the {\tt y} vector(s) in bytes\\
        \hline
        $cl$    & size of a cache line in bytes\\
        $b_{wT}$ & number of bytes written to memory  \\
        $b_{rT}$ & minimum number of bytes read from memory  \\
        $b_T$   & minimum number of bytes transferred  \\
        $B_{rT}$ & maximum number of bytes read from memory \\
        $B_T$   & maximum number of bytes transferred  \\
        $O$     & number of floating point operations \\
        $I_b$   & maximal computational intensity\\
        $I_w$   & minimum computational intensity\\        
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{Notations}
  \label{tab:not}
\end{figure}

@@

The matrix is composed of $n_r$ rows and columns with $n_z$ nonzeros
per row leading to a total of $$n_c = n_r n_z$$ nonzero entries in
the matrix. Each of these nonzero entries has one index of size $b_i$
and $n_m$ values of size $b_x$. The matrices have a total size
of $$s_M = n_c (b_i + b_x n_m) = n_r n_z (b_i + b_x n_m)$$

@@

On the other hand, in the worst case, every time the {\tt x} vector is
accessed, the value needs to be transfered from memory again. So in
total, there are as many transfers as the number of nonzeros in the
matrix. Note however that most architectures cannot read memory a
single byte at a time. Instead, a minimum number of bytes, equal to
the size of a cacheline $cl$, are transfered at once.  When there are
multiple vectors, each nonzero element uses $n_v$ consecutive
entries. The worst case number of bytes read and transfered are
$$B_{rT} = s_M + n_c cl \ceil{\frac{n_vb_x}{cl}} $$ 
$$B_T = n_v n_m b_x n_r + n_r n_z \left ( b_i + b_x n_m +  cl \ceil{\frac{n_vb_x}{cl}} \right)$$

In SpMV, each nonzero of the matrix requires two floating point
operations: one for performing the multiplication and one for
accumulating the result row-wise. Here we are dealing with $n_v n_m$
simultaneous SpMVs and the number of floating point operations is
$$O = 2 n_v n_m n_c = 2 n_v n_m n_z n_r$$

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show in Section~\ref{sec:expe}
that an implementation relying on compiler vectorization does not lead
to desirable performance, which confirms the results
of~\cite{Saule13-ARXIV}.) The registers in the MIC architecture are
512 bits wide and can store 16 floats or integers. (In comparison to
OpenCL or CUDA, one can think of these registers as a warp.) The end
goal is to be able to load and format the nonzero values $a,b,c,d$
and the vector entries $p,q,r,s$ in two vector registers to be able to
apply Fused Multiply-Add on them, which is depicted in
Figure~\ref{fig:tensor_product}. Bringing the data in the vector
register as depicted in the figure is the difficult part that we now
explain. A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
using an OpenMP construct.

@@

All the operations that loads data from the memory into a register
allows to transfer 512 bits, so 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we need to load at once the data for four nonzeros in the matrix,
representing 16 floats from the matrix and 16 floats from the vectors
at once. (The number of nonzeros on a row in our application is
always a multiple of 4, if it was, we would need to add some explicit
zeroes in the data structure to pad it.)

The values from the nonzero element in the matrix are the simplest to
load. One can load the 16 floating point values for four nonzeroes of
the matrix at once. The 4 floats coming from each matrix four 4
consecutive nonzeros are naturally grouped within each lane of the
SIMD vector. One can use a swizzle operation to broadcast the 4
elements of a single matrix to take the whole vector. (The broadcast of 
$A_h^1,A_h^2,A_h^3,A_h^4$ is shown in Figure~\ref{code:mat_mul}.)

Loading the entries from the {\tt x} vector is a little bit more
complicated. First the columns pointers of 4 different nonzeros are
loaded in a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to unpack for the
  LSB and MSB. Both are mandatory even if one is known not to be
  useful according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
replicated so that each fill a vector lane using a swizzle
operation. Then each value is multiplied by 4 (because there are 4
vectors) and the offsets $(3,2,1,0)$ is added each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a SIMD register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors accross
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize traffic on the memory bus.

@@

To perform our experiments and analysis, we use multiple kinds of
matrices which are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. That way we ensure the simplest form of multiplication,
the amount of elements from the {\tt x} vector data transfered from
the memory will be minimum. The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for most
application relying on sparse matrix vector multiplication. The Random
matrices (Figure~\ref{fig:compact}) see their nonzero element
randomly (uniformly) distributed in the matrix, they represent the
worst case scenario for cache based architecture where the cache
reutilisation is the lowest from one row to the next.

@@

In the previous sections, we have explored in the practical
implementation on a MIC architecture of multiple derivative operators
acting on multiple vectors within the context of RBF-RD. Each
derivative has an associated sparse matrix with a fixed number of
nonzeros on each row. While computing a single derivative of multiple
functions is rather common (vectorization occurs over the vectors), we
accelerate the algorithm further by considering multiple derivatives
with corresponding matrices with identical adjacency graphs. We
specialize to four matrices and four derivatives, with 16 outputs,
computed as a sum of outer products.
